Stagnating State Funding for Higher Education and its Effect on Faculty Composition at US Universities

### 2023-05-30 text/sections/06-discussion.tex
Finished round of writing revisions.  Only spell-checking remains.

The results show that the recent stagnation in state funding for higher education has affected the composition of faculty within public universities, away from tenure-track and tenured professors towards contingent lecturers.
At the same time, there are little discernible effects on individual professors hired in the years 2011-2021 at Illinois public universities, except for the salaries of first year lecturers.
The rate that faculty leave their university, and their rate of promotion between positions, are unaffected by the university revenues, so this leaves one primary channel to explain changes in faculty composition: hiring.

Between academic years the number of professors at a university changes: professors leave an institution (retiring, hired elsewhere, etc.) and other join (they are newly hired, end a sabbatical, etc.).
\autoref{sec:results-indiv} shows that most of these channels are unaffected, implying that falls in hiring for tenured-track and tenured faculty must explain most of the substitution towards lecturers away from tenured faculty.
Notably, faculty hiring is already know to be heavily impacted by budget shocks.
\cite{turner2014impact} documents the wide-spread practice of hiring freezes at universities in response to budget shocks around the 2008 recession.
Throughout the last decade, multiple such measures were taken by Illinois public universities in response to their deteriorating finances \citep{furlough2010}.
The University of Illinois\footnote{
    The University of Illinois includes three public university campuses: Urbana-Champaign, Chicago, Springfield.
}
did not receive the allocated state appropriations from the state of Illinois on time, so enacted cost-cutting measures to stay fiscally solvent.
In response, the university system placed a hold on all hiring for filling state-funded positions and promotions.

The data used so far measure faculty count, so cannot disentangle which of these channels is most affected by changes in state funding at a public university, yet \cite{wapman2022quantifying} provide a measure of the aggregated total faculty hires at universities over the time period 2011--2021.\footnote{
    \citep{wapman2022quantifying} made an aggregated version of their data open-source, making this sub-analysis possible.
}
\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{State Funding and Faculty Hired at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.8\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation.png}
\end{figure}

Universities with more state funding (for the entire decade) hired more professors (\autoref{fig:hiring-correlation.png}).
Similarly, the funding shock instrument shows that an increase in 10\% state funding leads to 13\% more faculty hires at a university across the decade 2011--2021 (see \autoref{tab:hiring-shock-reg}).\footnote{
    These results were produced by integrating the total count of faculty hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
    There were no observable differences in the hiring rate of male vs female faculty.
    See \autoref{sec:appendix-hiring} for further details.
}
It is clear that public universities' hiring is negatively impacted by falls in state funding, even using highly aggregated data for a total count of hiring across an entire decade.
While this analysis strengthens the case negative effects on hiring plays, it is crude for measuring exact magnitudes for the effect on faculty hiring.
I leave it to further research to delve deeper into the magnitudes that hiring plays as the primary causal mediator for the effects of falls in state funding on faculty composition at public universities.

>>>>

The results show that the recent stagnation in state funding for higher education has affected the composition of faculty within public universities, away from tenure-track and tenured professors towards contingent lecturers.
At the same time, there are little discernible effects on individual professors hired in the years 2011-2021 at Illinois public universities.
The rate that faculty leave their university, and their rate of promotion between positions, are unaffected by the university revenues, so this leaves one primary channel to explain changes in faculty composition: hiring.

The number of professors at a university changes between academic years: professors leave an institution (retiring, hired elsewhere, etc.), while others join (they are newly hired, end a sabbatical, etc.).
Section~\ref{sec:results-indiv} shows that most of these channels are unaffected, implying that falls in hiring for tenured-track and tenured faculty must explain most of the substitution towards lecturers and away from tenured professors.
Notably, faculty hiring is suspected to be impacted by changes in university funding; for example, \cite{turner2014impact} documents the wide-spread practice of hiring freezes at universities in response to budget shocks around the 2008 recession.
Similar measures were taken by Illinois public universities in response to their deteriorating finances in the 2010s \citep{furlough2010}.
The University of Illinois did not receive funding from the state of Illinois before the next fiscal year started, so enacted cost-cutting measures to stay fiscally solvent in the interim.
As one such cost-cutting measure, the university system placed a freeze on all hiring for filling state-funded positions and promotions.

The data used so far measure faculty count, so cannot disentangle which channels are most affected by changes in state funding at a public university.
\autoref{fig:hiring-correlation.png} uses data on the total faculty hired at US public universities over 2011--2021, provided by \cite{wapman2022quantifying},\footnote{
    \cite{wapman2022quantifying} provided an aggregated sample of their data as open-source, making this analysis possible.
} showing a correlation between professor hiring rate and the total state funding per student.
\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{State Funding and Faculty Hired at Public Universities, Total for 2011--2021.}
    \includegraphics[width=0.8\textwidth]{figures/hiring-correlation.png}
    \label{fig:hiring-correlation.png}
\end{figure}

Universities with more state funding (for the entire decade) hired more professors (\autoref{fig:hiring-correlation.png}).
Similarly, the funding shock IV model shows that a decrease of $-10$\% in state funding leads to $-13$\% fewer faculty hires per student across the decade 2011--2021 (see \autoref{tab:hiring-shock-reg} for full regression model results).\footnote{
    These results were produced by integrating the total count of faculty hires for 2010--2021 for the top-ranked 180 US universities with a sum of the funding variables, and then estimating the models specified in \autoref{sec:iv-model-uni}.
    There were no observable differences in the hiring rate of male vs female faculty.
    See \autoref{sec:appendix-hiring} for further details.
}
It is clear that public universities' hiring is negatively impacted by falls in state funding, even using this highly aggregated count of professor hiring across an entire decade.
While this analysis strengthens the case negative effects on hiring plays, it is crude for measuring exact magnitudes for the effect on faculty hiring.
I leave it to further research to delve deeper into the magnitudes that hiring plays as the primary causal mediator for the effects of falls in state funding on faculty composition at public universities.

### 2023-05-30 text/sections/07-conclusion.tex
Finished round of writing revisions.  Only spell-checking remains.

Findings show that public universities have substituted away from tenure-track and tenured professors, towards non-tenured lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, which implies that the effect is driven by changes in university hiring of new professors.
Each year, public universities are using more contingent lecturers to teach their education students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts.
Private universities have always had more tenured professor per student than public institutions, but each year the gap widens.

While costs education have been rising in the US, thanks to multiple overlapping trends \citep{ehrenberg2012}, public universities have also dealt with declining state support.
It is natural to expect that such headwinds will lead to systematic change at public universities, changes which are not neutral to the composition of faculty, or to their goals of research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
If public universities' hiring of faculty is disrupted by a fall in state funding, then we can expect that private universities will benefit (in a competitive sense) by being able to hire better professors from relatively resource constrained public universities.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects that restricting their funding has on the composition of faculty, and the impact on higher education as a whole.

>>>>

Public universities have systematically substituted away from tenure-track and tenured professors, towards non-tenured lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, while suggestive information shows falls in hiring, which together implies that the changes in faculty composition are driven by reduced hiring of new professors.
Each year, public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

While costs education have been rising in the US, public universities have also dealt with declining state funding.
It is natural to expect that such headwinds will lead to systematic change at public universities, changes that affect their faculty, and limit the universities' goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty composition, research at public universities, and the impact on higher education as a whole.
